Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-07 source/paper.tex
some description of the timestepping algorithms

\JScomment{This paragraph should refer to the schematic and explain
  all aspects} The Gusto model comprises a domain and a prognostic
equation set. The domain holds the mesh and the finite element
function spaces along with any other information about the space-time
domain such as the coordinates and the timestep. The prognostic
equation sets are each defined in a python class, making use of the
common finite element forms for the transport and diffusion terms to
set up the residual (i.e. the left hand side of the equations defined
in section \ref{sec: equations}) in weak form. At this point, no
particular discretisation of the transport or diffusion forms is
specified. This happens in the \texttt{SpatialMethod} class which can
be used to apply the methods described in sections \ref{sec:
  transport} and \ref{sec: diffusion} to the relevant terms. Any
physics source terms, such as those required for moist processes or
tubulence parameterisation, are added by calling the \texttt{Physics}
class that defines them. The \texttt{TimeStepper} class contains the
definition of the timestepping procedure and the \texttt{run} method
that loops over the timesteps. It also sets up diagnostic fields and
calls the IO as required. The simplest timeloop class will just apply
one time discretisation method to the equation set but it is common in
weather and climate prediction models to split the different types of
terms and apply a time discretisation separately to each type of
term. For example, the transport and forcing terms might be solved for
separately, such as in **, which enables explicit transport schemes to
be applied to each field. ** something about SIQN structure. Terms
involved in physics parameterisations are evaluated in the appropriate
part of the iteration loop, with physics processes that occur on
`fast' timescales evaluated more frequently and `slower' physics
processes evaluated just once at the end of the timestep. Since the
test cases presented later only use simple moist physics
parameterisations, we just include the option to solve for these at
the end of the timestep.

\begin{algorithm} 
  \caption{Basic timestepping scheme}
  \label{alg: basic}
  \begin{algorithmic}[1]
    \STATE Replace transport and diffusion terms with specified spatial discretisation scheme.
    \STATE Setup IO and diagnostics and write out initial conditions.
    \WHILE{$t < t_{max}$}
      \STATE Apply timestepping scheme to equation set: $\chi^{n+1} = \mathcal{T}(\chi^n, \chi^{n+1})$
      \STATE Advance timestep: $\chi^n = \chi^{n+1}$
      \IF{ndt \% nts == 0}
        \STATE Update timeseries diagnostics file.
      \ENDIF
      \IF{ndt \% ndiag == 0}
        \STATE Compute and write out diagnostic fields.
      \ENDIF
      \IF{ndt \% nchk == 0}
        \STATE Write out checkpoint file.
      \ENDIF
    \ENDWHILE
  \end{algorithmic} 
\end{algorithm}

>>>>

\JScomment{This paragraph should refer to the schematic and explain
  all aspects} The Gusto model comprises a domain and a prognostic
equation set. The domain holds the mesh and the finite element
function spaces along with any other information about the space-time
domain such as the coordinates and the timestep. The prognostic
equation sets are each defined in a python class, making use of the
common finite element forms for the transport and diffusion terms to
set up the residual (i.e. the left hand side of the equations defined
in section \ref{sec: equations}) in weak form. At this point, no
particular discretisation of the transport or diffusion forms is
specified. This happens in the \texttt{SpatialMethod} class which can
be used to apply the methods described in sections \ref{sec:
  transport} and \ref{sec: diffusion} to the relevant terms. Any
physics source terms, such as those required for moist processes or
tubulence parameterisation, are added by calling the \texttt{Physics}
class that defines them. The \texttt{TimeStepper} class contains the
definition of the timestepping procedure and the \texttt{run} method
that loops over the timesteps. It also sets up diagnostic fields and
calls the IO as required. The simplest timeloop class will just apply
one time discretisation method to the equation set but it is common in
weather and climate prediction models to split the different types of
terms and apply a time discretisation separately to each type of
term. For example, in \citet{} the transport and forcing terms are
solved separately, enabling subcycling of transport schemes and a
larger stable Courant number. Terms involved in physics
parameterisations are evaluated in the appropriate part of the
iteration loop, with physics processes that occur on `fast' timescales
evaluated more frequently and `slower' physics processes evaluated
just once at the end of the timestep. Since the test cases presented
later only use simple moist physics parameterisations, we only
describe the option to solve for these at the end of the timestep.

In Algorithm \ref{alg: basic} we show the basic structure of the
timestepping loop in Gusto. The first step is to modify the equation
residual to apply the spatial discretisation (for example, the DG
upwind or SUPG discretisations described in section \ref{sec:
  transport}). Before entering the timestepping loop, we set up the
IO, including any diagnostics, and write out the initial
conditions. In this basic timestepper, the timestepping loops just
applies the timestepping scheme to the equation set. Any output is
then decised based on the frequencies specified by the user. For
example, we output a timeseries of diagnostics, such as minimum and
maximum values of the fields, every \texttt{nts} timesteps; we output
the diagnostic fields every \texttt{ndiag} timesteps and we write out
checkpointing files every \texttt{nchk} timesteps. Checkpointing files
can be used to initialise a new simulation or to compare to previous
output.

Algorithms \ref{alg: split physics} and \ref{alg: SIQN} we illustrate
two more complex timestepping algorithms. These replace lines 4-5 in
Algorithm \ref{alg: basic}.

\begin{algorithm} 
  \caption{Basic timestepping scheme}
  \label{alg: basic}
  \begin{algorithmic}[1]
    \STATE Replace transport and diffusion terms with specified spatial discretisation scheme.
    \STATE Setup IO and diagnostics and write out initial conditions.
    \WHILE{$t < t_{max}$}
      \STATE Apply timestepping scheme to equation set: $\chi^{n+1} = \mathcal{T}(\chi^n, \chi^{n+1})$
      \STATE Advance timestep: $\chi^n = \chi^{n+1}$
      \IF{ndt \% nts == 0}
        \STATE Update timeseries diagnostics file.
      \ENDIF
      \IF{ndt \% ndiag == 0}
        \STATE Compute and write out diagnostic fields.
      \ENDIF
      \IF{ndt \% nchk == 0}
        \STATE Write out checkpoint file.
      \ENDIF
      \STATE $t \gets t + \Delta t$
    \ENDWHILE
  \end{algorithmic} 
\end{algorithm}
